Lossy compression of DNA sequencing quality data

### 2015-05-01 manuscript/q-compression.tex
More revisions to the text, and fixed an error in a reference entry.

As large sets of quality strings show similar trends of quality over
their length, it may be possible to identify such common patterns in
the data and use them as reference profiles to approximate individual
sequences of quality values. Here we use $k$-means clustering, a
vector quantization method, which partitions a set of samples into $k$
sets that minimize within-cluster sum of
squares~\citep{macqueen1967some}. We sampled 10K reads at random and
computed cluster centers as read quality profiles using heuristic
iterative refinement approach by quickly converging to a locally
optimal minimum~\citep{hartigan1979algorithm}.  The obtained cluster
were then used quality profile prototypes for the dataset.  All reads
are evaluated with the nearest quality profile in Euclidean space
being assigned to every read as their compressed representation.  The
compressed quality file therefore consists of an index enumerating the
$k$ quality profiles, and a binary part containing the assigned
quality profile index for each read.  Although this approach is not
able to capture some read specific differences in quality values, it
does approximates the overall patterns in quality values. An example
of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the Supplemental Table 1. Storing the uncompressed quality values
requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in
the figure. The lossless compression of each data set using
\textsc{bz}ip2 ranges from 2.19 - 3.10 bits/base-call and is denoted
by the colored asterisks on the abscissa. The compression methods tend
to cluster together across the different data sets. Across all data
sets, the 0-degree polynomial regression, profile encodings, and
\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires 10-15x more
storage than the profile encoding methods for only a 2-3x reduction in
\textsc{mse}. When \textsc{q}ual\textsc{c}omp's rate parameter is set
to match the profile encoding methods, \textsc{q}ual\textsc{c}omp
performs slightly worse in terms of \textsc{mse}. In the
\textit{Rhodobacter sphaeroides} 2.4.1 fragment library,
\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has
a \textsc{mse} of 17.29. Using 256-profile encoding requires less
storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set was
best in all metrics (Table \ref{fig:assembly_ranks}). Among the
compression methods, the profile encoding performed best, polynomial
regression modeling performed worst, and other methods were
intermediate (Fig.~\ref{fig:assembly_ranks}).

@@

It is important to highlight the result that the 2-bin method aligns a
greater portion of reads than the various regression models. If a
poor quality base is flanked by high quality sequence, then the low
degree polynomial regression models tend to smooth out the base
quality, erroneously marking the low quality base as higher quality.
During alignment, Bowtie2 penalizes mismatches at high quality bases
more so than mismatches at lower quality bases. Thus, the polynomial
regression models incur a high penalty for these bases, resulting in
less alignments, despite having a better \textsc{mse} than 2-bin in
most cases.

@@

It is important to highlight that the downstream applications dictate
which lossy compression methods are acceptable. As highlighted
earlier, 2-bin encoding is able to retain information about low
quality bases that occur within stretches of high quality bases,
whereas the polynomial regression methods tend to smooth these
qualities.  In addition to aligning less reads that the 2-bin
encoding, the polynomial regression methods may also be hindering
downstream tools that use the alignment results. A \textsc{snp}
detection pipeline will commonly use the quality values to record if a
potential difference from the reference is the result of genetic
variation or sequencing error.  Erroneously marking a base as higher
quality, may lead to the base being incorrectly labeled as a
\textsc{snp}.  2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly. In the case of assembly,
the 7th-degree polynomial regression method outperforms 2-bin encoding
in nearly all metrics.

>>>>

Sets of quality strings show similar trends of quality over their
length, and it is possible to identify common patterns in the data and
use them as reference profiles to approximate individual sequences of
quality values. Here we use $k$-means clustering, a vector
quantization method that partitions a set of samples into $k$ sets
that minimize within-cluster sum of
squares~\citep{macqueen1967some}. We sampled 10$K$ reads at random and
computed cluster centers as read quality profiles using heuristic
iterative refinement approach by quickly converging to a locally
optimal minimum~\citep{hartigan1979algorithm}.  All reads are
evaluated with the nearest quality profile in Euclidean space being
assigned to every read as their compressed representation.  The
compressed quality file therefore consists of an index enumerating the
$k$ quality profiles, and a binary part containing the assigned
quality profile index for each read.  Although this approach is not
able to capture some read specific differences in quality values, it
does approximates the overall patterns in quality values. An example
of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides}, and \textit{Homo sapiens} data
sets, but the additional short-jump library results are available in
the Supplemental Table 1. Storing the uncompressed quality values
requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in
the figure. The lossless compression of each data set using
\textsc{bz}ip2 ranges from 2.19 - 3.10 bits/base-call and is denoted
by the colored asterisks on the abscissa. The compression methods tend
to cluster together across the different data sets. Across all data
sets, the 0-degree polynomial regression, profile encodings, and
\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires 10-15x more
storage than the profile encoding methods for only a 2-3x reduction in
\textsc{mse}. When \textsc{q}ual\textsc{c}omp's rate parameter is set
to match the profile encoding methods, \textsc{q}ual\textsc{c}omp
performs slightly worse in terms of \textsc{mse}. In the
\textit{Rhodobacter sphaeroides} fragment library,
\textsc{q}ual\textsc{c}omp with rate 10 bits/read (0.099 bits/bp) has
a \textsc{mse} of 17.29. Using 256-profile encoding requires less
storage (0.079 bits/bp) and has a lower \textsc{mse} (11.85).

@@

No assembly of the \textit{Rhodobacter sphaeroides} data set was
best in all metrics (Table \ref{fig:assembly_ranks}). Among the
compression methods, the profile encoding performed best, polynomial
regression modeling performed worst, and other methods were
intermediate (Fig.~\ref{fig:assembly_ranks}).

@@

Note that the 2-bin method aligns a
greater portion of reads than the various regression models. If a
poor quality base is flanked by high quality sequence, then the low
degree polynomial regression models tend to smooth out the base
quality, erroneously marking the low quality base as higher quality.
During alignment, Bowtie2 penalizes mismatches at high quality bases
more so than mismatches at lower quality bases. Thus, the polynomial
regression models incur a high penalty for these bases, resulting in
less alignments, despite having a better \textsc{mse} than 2-bin in
most cases.

@@

It is important to note that the downstream applications dictate
which lossy compression methods are acceptable. As noted
above, 2-bin encoding is able to retain information about low
quality bases that occur within stretches of high quality bases,
whereas the polynomial regression methods tend to smooth these
qualities. In addition to aligning less reads that the 2-bin
encoding, the polynomial regression methods may also be hindering
downstream tools that use the alignment results. A \textsc{snp}
detection pipeline will commonly use the quality values to record if a
potential difference from the reference is the result of genetic
variation or sequencing error.  Erroneously marking a base as higher
quality, may lead to the base being incorrectly labeled as a
\textsc{snp}.  2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly. In the case of assembly,
the 7th-degree polynomial regression method outperforms 2-bin encoding
in nearly all metrics.
